Retrofitting Effect Handlers onto OCaml

### 2020-11-17 retro-concurrency.tex
edits

Similarly to~\cite{Leijen14, Kiselyov12}, Multicore OCaml uses the call stack for
implementing continuations, but our continuations are one-shot. Bruggeman et
al.~\cite{Bruggeman96} show how to implement one-shot continuations efficiently
using segmented stacks in Scheme. Farvardin et al.~\cite{Farvardin20} perform a
comprehensive evaluation of various implementation strategies for continuations
on modern hardware. Multicore OCaml stacks do not neatly fit the description of
one of these implementation strategies -- they are best described as using the resize
strategy from Farvardin et al. for each of the fibers, which are in turn linked
to represent the current stack and the captured continuations.

@@

The interaction of non-local control flow and resources has been studied
extensively. Scheme uses |dynamic-wind|~\cite{R5RS}, which is a generalisation of
Common Lisp |unwind-protect|~\cite{Steele90}, which ensures de-allocation (and
re-allocation) of resources every time the non-local control leaves and enters
back into a context. |dynamic-wind| is not quite sufficient for resource
clean up~\cite{Kiselyov,Sitaram03} as returning exits must be distinguished
from non-returning ones, and resource cleanup is only necessary during
non-returning exits.

In Multicore OCaml, we build on the existing defensive coding practices against
exceptions to clean up resources on non-returning exits. Our continuations are
assumed to be resumed exactly once using |continue| or |discontinue|. As a
result, when a computation performs an effect, we are guaranteed that the
control will return. For the non-returning cases (value and exceptional
return), the code already handles resource cleanup. OCaml does not have a
\texttt{\footnotesize try/finally} construct commonly used for resource cleanup
in many programming languages. Alternative standard libraries such as
Base~\cite{BaseProtect} and Core~\cite{CoreProtect} provide mechanisms
analogous to |unwind-protect|, which are in turn implemented using exceptions.
Thus, the linear use of continuations enabled by the |discontinue| primitive
ensures backwards compatibility of legacy OCaml systems code under non-local
control flow introduced by effect handlers. Leijen~\cite{Leijen18} explicitly
extends effect handlers with |initially| and |finally| clauses in Koka for
resource safety.

@@

We presented the motivation, design and implementation of effect handlers
specifically aimed at retrofitting them onto OCaml with tool compatibility,
negligible overheads for existing OCaml programs and good performance for those
that use effect handlers. Our evaluation shows that this is indeed the case.

>>>>

Multicore OCaml uses the call stack for implementing continuations (as do~\cite{Leijen14, Kiselyov12}), but with one-shot continuations. Bruggeman et
al.~\cite{Bruggeman96} show how to implement one-shot continuations efficiently
using segmented stacks in Scheme. Farvardin et al.~\cite{Farvardin20} perform a
comprehensive evaluation of various implementation strategies for continuations
on modern hardware. Multicore OCaml stacks do not neatly fit the description of
one of these implementation strategies -- they are best described as using the resize
strategy from Farvardin et al. for each of the fibers, which are in turn linked
to represent the current stack and the captured continuations.

@@

The interaction of non-local control flow and resources has been studied
extensively. Scheme uses |dynamic-wind|~\cite{R5RS}, which is a generalisation of
Common Lisp |unwind-protect|~\cite{Steele90}, which ensures de-allocation (and
re-allocation) of resources every time the non-local control leaves and enters
back into a context. |dynamic-wind| is not quite sufficient for resource
clean up~\cite{Kiselyov,Sitaram03} as returning exits must be distinguished
from non-returning ones, and resource cleanup is only necessary during
non-returning exits.
Multicore OCaml builds on the existing defensive coding practices against
exceptions to clean up resources on non-returning exits. Our continuations are
assumed to be resumed exactly once using |continue| or |discontinue|. As a
result, when a computation performs an effect, we are guaranteed that the
control will return. For the non-returning cases (value and exceptional
return), the code already handles resource cleanup. OCaml does not have a
\texttt{\footnotesize try/finally} construct commonly used for resource cleanup
in many programming languages. Alternative standard libraries such as
Base~\cite{BaseProtect} and Core~\cite{CoreProtect} provide mechanisms
analogous to |unwind-protect|, which are in turn implemented using exceptions.
Thus, the linear use of continuations enabled by the |discontinue| primitive
ensures backwards compatibility of legacy OCaml systems code under non-local
control flow introduced by effect handlers. Leijen~\cite{Leijen18} explicitly
extends effect handlers with |initially| and |finally| clauses in Koka for
resource safety.

@@

We have presented the design and implementation of effect handlers
aimed at retrofitting them onto the large body of existing OCaml
code.  Our design walks the tightrope of maintaining
compatibility (e.g. for profiling and debuggers and negligible overheads for existing 
programs) and also demonstrating good performance and usability for new OCaml
code that takes advantage of effect handlers. Our implementation in Multicore OCaml
deviates from previous categorisations~\cite{Farvardin20} and our evaluation shows
that we have achieved our goal of balancing compatibility with the introduction
of effect handlers for mainstream OCaml programming.
